# Advanced R

## Machine learning

## Deep learning

## Text mining


### Sentiment Analysis 

Sentiment Analysis (SA) extracts information on emotion or opinion from natural language (Silge and Robinson 2017). Most forms of SA provides information about positive or negative polarity, e.g. whether a tweet is positive or negative.


This tutorial leverages the data provided in the janeaustenr package. This package contains the complete text of Jane Austen's 6 completed, published novels, formatted to be convenient for text analysis.


The tidytext package contains three sentiment lexicons in the sentiments dataset.


The three lexicons are

  - AFINN from Finn Årup Nielsen
  - bing from Bing Liu and collaborators
  - nrc from Saif Mohammad and Peter Turney

```{r}
library(tidytext)
library(janeaustenr)
library(dplyr)
library(stringr)
library(ggpubr)
library(textdata)

tidy_books <- austen_books() %>%
  group_by(book) %>%
  mutate(
    linenumber = row_number(),
    chapter = cumsum(str_detect(text, 
                                regex("^chapter [\\divxlc]", 
                                      ignore_case = TRUE)))) %>%
  ungroup() %>%
  unnest_tokens(word, text)
```


```{r}
tidy_books %>% group_by(book) %>% count()
```

```{r}
nrc_se <- get_sentiments("nrc")
```

```{r}
nrc_se %>% group_by(sentiment) %>% count()
```


```{r}
nrc_bing <- get_sentiments("bing")
```

```{r}
nrc_bing %>% group_by(sentiment) %>% count()
```


```{r}
nrc_afinn <- get_sentiments("afinn")
```

```{r}
nrc_afinn %>% group_by(value) %>% count()
```






```{r}
nrc_joy <- get_sentiments("nrc") %>% 
  filter(sentiment == "joy")
tidy_books %>%
  filter(book == "Emma") %>%
  inner_join(nrc_joy) %>%
  count(word, sort = TRUE)
```



```{r}
library(tidyr)
jane_austen_sentiment <- tidy_books %>%
  inner_join(get_sentiments("bing")) %>%
  count(book, index = linenumber %/% 80, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>%
  mutate(sentiment = positive - negative)
```



create an index that breaks up each book by 500 words; this is the approximate number of words on every two pages so this will allow us to assess changes in sentiment even within chapters


```{r}
library(ggplot2)

ggplot(jane_austen_sentiment, aes(index, sentiment, fill = book)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~book, ncol = 2, scales = "free_x") 
```


```{r}
pride_prejudice <- tidy_books %>% 
  filter(book == "Pride & Prejudice")

pride_prejudice

```


```{r}
afinn <- pride_prejudice %>% 
  inner_join(get_sentiments("afinn")) %>% 
  group_by(index = linenumber %/% 80) %>% 
  summarise(sentiment = sum(value)) %>% 
  mutate(method = "AFINN")

bing_and_nrc <- bind_rows(
  pride_prejudice %>% 
    inner_join(get_sentiments("bing")) %>%
    mutate(method = "Bing et al."),
  pride_prejudice %>% 
    inner_join(get_sentiments("nrc") %>% 
                 filter(sentiment %in% c("positive", 
                                         "negative"))
    ) %>%
    mutate(method = "NRC")) %>%
  count(method, index = linenumber %/% 80, sentiment) %>%
  pivot_wider(names_from = sentiment,
              values_from = n,
              values_fill = 0) %>% 
  mutate(sentiment = positive - negative)
```



```{r}
bind_rows(afinn, 
          bing_and_nrc) %>%
  ggplot(aes(index, sentiment, fill = method)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~method, ncol = 1, scales = "free_y")
```


```{r}
get_sentiments("nrc") %>% 
  filter(sentiment %in% c("positive", "negative")) %>% 
  count(sentiment)
```


```{r}
get_sentiments("bing") %>% 
  count(sentiment)
```


```{r}
bing_word_counts <- tidy_books %>%
  inner_join(get_sentiments("bing")) %>%
  count(word, sentiment, sort = TRUE) %>%
  ungroup()
```


```{r}
bing_word_counts %>%
  group_by(sentiment) %>%
  slice_max(n, n = 10) %>% 
  ungroup() %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(n, word, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, scales = "free_y") +
  labs(x = "Contribution to sentiment",
       y = NULL) +
  theme_pubr()
```


```{r}
custom_stop_words <- bind_rows(tibble(word = c("miss"),  
                                      lexicon = c("custom")), 
                               stop_words)
```



```{r}
library(wordcloud)
tidy_books %>%
  anti_join(stop_words) %>%
  count(word) %>%
  with(wordcloud(word, n, max.words = 100))
```


```{r}
library(reshape2)

tidy_books %>%
  inner_join(get_sentiments("bing")) %>%
  count(word, sentiment, sort = TRUE) %>%
  acast(word ~ sentiment, value.var = "n", fill = 0) %>%
  comparison.cloud(colors = c("gray20", "gray80"),
                   max.words = 100)
```



